---
title: "624. HW2. Fall 2019"
author: "Mikhail Groysman"
date: "9/20/2019"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 5

---

Exercise come from book "Forecasting: Principles and Practice" by Rob J Hyndman and George Athanasopoulos, Chapter 6 "Time series decomposition"

# 6.2

The plastics data set consists of the monthly sales (in thousands) of product A for a plastics manufacturer for five years.

a. Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?
b. Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.
c. Do the results support the graphical interpretation from part a?
d. Compute and plot the seasonally adjusted data.
e. Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?
f. Does it make any difference if the outlier is near the end rather than in the middle of the time series?

## A

First, we will load plastics data.

```{r, message=FALSE, warning=FALSE}

#install.packages("forecast")

#install.packages("fma")

library(forecast)

library(ggplot2)

library(fma)

plastics <- fma::plastics

head(plastics)

summary(plastics)

```


We can see that no values is missing and data ranges from 697 to 1637.

Let's plot our data.

```{r}

autoplot(plastics) +
  ggtitle("Monthly sales of plastics product A - 5 year time period") +
  xlab("Montly") +
  ylab("In thousands")

```

Data is heavily seasonal.Dropping in winter and peaking in summer. Overall trend is positive.

## B

Let's do multuplicative decomposition

```{r}

plastics_decomp<-decompose(plastics,type="multiplicative")

head(plastics_decomp)

summary(plastics_decomp)


```

Let's plot decomposed data:

```{r, results='hide',warning=FALSE}

autoplot(plastics_decomp) +
  ggtitle("Monthly sales of plastics product A - 5 year time period - Multiplicative Decomposition") +
  xlab("Montly") 
```

## C

The decomposition shows strong seasonality in middle of year(summer), and overall positive trend with a small dip in year 5. Except for dip in year 5, visual interpretation from (a) was consistent with decomposition.

## D 

Let's do calculate and plot the sesonally adjusted data.

```{r}

#install.packages("seasonal")

library(seasonal)

adjust_plastics<-plastics/plastics_decomp$seasonal

autoplot(adjust_plastics) +
  ggtitle("Monthly sales of plastics product A - 5 year time period - Seasonally Adjusted") +
  xlab("Montly") +
  ylab("In thousands")


```

Seasonal Adjustment shows that the first year sales were flat and then they grew for the next 3 years. The 5th year saw a drop in sales. 

## E

Let's make the 1st observation 500 higher.

```{r}

plastics[1]<-plastics[1]+500

head(plastics)

```

Let's recalculate seasonaly adjusted data.


```{r}

plastics_decomp<-decompose(plastics,type="multiplicative")

adjust_plastics<-plastics/plastics_decomp$seasonal

autoplot(adjust_plastics) +
  ggtitle("Monthly sales of plastics product A - 5 year time period - Seasonally Adjusted") +
  xlab("Montly") +
  ylab("In thousands")

```

As expected, we see a big drop in the beginning of the 1st year, then data is flat. The data grows for the next 3 year, and then drops in year 5. I do not see much an effect.


## F

Let's try both scenarios.

In the middle, first

```{r}

plastics[1]<-plastics[1]-500

plastics_m<-plastics

plastics_m[30]<-plastics_m[30]+500

```

Let's recalculate seasonaly adjusted data and plot it.

```{r}

plastics_decomp_m<-decompose(plastics_m,type="multiplicative")

adjust_plastics_m<-plastics_m/plastics_decomp_m$seasonal

autoplot(adjust_plastics_m) +
  ggtitle("Monthly sales of plastics product A - 5 year time period - Seasonally Adjusted") +
  xlab("Montly") +
  ylab("In thousands")

```

We see data our data spike in the middle, as we expected. Also, overall data looks less smooth.

Next, let's try to put an outlier in the end of our data.

```{r}

plastics_e<-plastics

plastics_e[54]<-plastics_e[54]+500

```

Let's recalculate seasonaly adjusted data and plot it.

```{r}

plastics_decomp_e<-decompose(plastics_e,type="multiplicative")

adjust_plastics_e<-plastics_e/plastics_decomp_e$seasonal

autoplot(adjust_plastics_e) +
  ggtitle("Monthly sales of plastics product A - 5 year time period - Seasonally Adjusted") +
  xlab("Montly") +
  ylab("In thousands")

```

Now, data seems to increase with a big spike in the beginning of the middle of year 5, then a drop. Again, data appears to be less smooth.

I see that putting an outlier in the end, makes a drop in sales shorter in duration. 
